Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-07-13 ms.tex
Removing undefined reference, rewording Neufeld solution description
Finished first draft?

\citet{1973MNRAS.162...43H, 1974MNRAS.166..373H} demonstrated that resonance-line radiation transfer in low-density, optically-thick media can be described by the Poisson equation. However, the accuracy of their solution is limited by the treatment of the boundary conditions. In their eigenfunction expansion, the spatial and frequency components of the differential equation are separated, yet the spatial eigenvalues are defined as being explicitly frequency-dependent. The resulting solution is therefore invalid on the basis that the differential equation and boundary conditions cannot simultaneously be satisfied. A more detailed discussion of the boundary condition is included in Section \ref{sec:steadystate}. \citet{1990ApJ...350..216N} follows this approach in an attempt to extend the result to media of intermediate optical depth, including the effects of scattering in the Doppler core of the line, but again does not use the correct boundary condition. \citet{?} improved on this approach, producing a divergent, inhomogeneous solution to the problem in 1D slab geometry that includes the effect of photon absorption within the medium. \citet{2006ApJ...649...14D} generalize the problem to spherical geometry, which is the same physical set up used here. Following this, \citet{2020arXiv200509692L} presents closed-form solutions for the radiation field in spherical geometry, utilizing the ``core skipping'' method discussed in \citet{2015MNRAS.449.4336S} to handle photon frequency redistribution within the Doppler core. This technique involves drawing the atom's velocity from a truncated Gaussian distribution, forcing photons back into the wing when they scatter into the core. This is motivated by the assumption that all core scattering results in zero spatial diffusion until the photon is redistributed back out into the wing where the mean free path becomes much larger and it can diffuse in space again. However, it is not clear whether this prescription is fully independent of the choice of velocity distribution.

@@

The diffusion equation for photons escaping a uniform sphere of gas is a function of time, which was ignored in the steady-state calculations in Section \ref{sec:steadystate}. To obtain the radiative intensity $I=dE/(dAdtd\Omega d\nu)$ of a fluid that is dynamic on timescales shorter than the light-crossing time $t_{\rm lc} = R/c$, the time-dependent response to an impulse is found. This allows the distribution of photon escape times (the ``wait time distribution'') to be characterized and fit for efficient sampling. For simplicity a $J=0$ boundary condition will be used in the following derivations, which is expected to be a decent approximation for $\tau_0 \gg 1$.

@@

Let us define the damping time to be $\gamma \equiv i\omega$, which is real-valued and positive for damped solutions. There are certain values of $\gamma$ for which a resonant response is produced which we will call $\gamma_{nm}$. Here, $m$ is an index that represents individual frequency eigenmodes, as opposed to $n$ which represents the spatial mode number. Near these frequencies, the solution will have the form 
\be
J(n,\sigma,-i\gamma) & \simeq \frac{ J_{nm}(\sigma) }{\gamma_{nm} - \gamma} + {\rm terms\ which\ vary\ slowly\ in\ \gamma}.
\ee
Obtaining the values of these eigenfrequencies and their corresponding eigenfunctions $J_{nm}(\sigma)$ allows us to obtain the contribution of this mode to the total solution via contour integration. 
\be
J(r,\sigma,t) & = &  -j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t}J(n,\sigma,-i\gamma)
\nonumber \\ & \simeq & j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t} \frac{ J_{nm}(\sigma) }{\gamma - \gamma_{nm}} 
\nonumber \\ & \simeq & j_0(\kappa_n r)  J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Here we have closed the integral along the imaginary $\gamma$ axis by a semi-circle at infinity, using the residue theorem to obtain the answer. The contribution from the semi-circle at infinity is zero. Summing over all spatial eigenmodes $n$ and over all temporal eigenmodes $m$ for a given $n$, we get the final form of the result
\be \label{eq:Jrsigmat}
J(r,\sigma,t) & = & \sum_{n=1}^\infty j_0(\kappa_n r)  \sum_m J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Beginning with Eq. \ref{eq:Jrsigmat}, we use the derivative
\be
\frac{dj_0(\kappa_n r)}{dr} \rfloor_R & =& \frac{d}{dr}  \rfloor_R \left( \frac{\sin(\kappa_n r)}{\kappa_n r} \right)
=  \left( \frac{\cos(\kappa_n R)}{R} - \frac{\sin(\kappa_n R)}{\kappa_n R^2} \right)\rfloor_R
\nonumber \\ & = & \frac{(-1)^n}{R},
\ee
to obtain the flux at the surface, which is
\be
F(R,\sigma,t) & =& - \frac{4\pi}{3k\phi} \frac{dJ(R,\sigma,t)}{dr} 
= - \frac{4\pi}{3k\phi R}  \sum_{nm} (-1)^n J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Multiplying by $4\pi R^2$ gives the energy per time per frequency
\be
\frac{dE}{dtd\nu} & = & - \frac{16\pi^2 R}{3k\phi}  \sum_{nm} (-1)^n J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\label{eq:dEdtdnu}
\ee
Integrating this over time yields a factor $-1/\gamma_{nm}$, and over $d\nu = \sqrt{3/2} \Delta^2 \phi d\sigma$, a ``sum rule" results
\be
1 & = &  \sqrt{ \frac{3}{2} } \frac{16\pi^2R\Delta^2}{3kE} \sum_{nm} (-1)^n \gamma_{nm}^{-1} \int d\sigma J_{nm}(\sigma).
\ee
This non-trivial expression provides a check on the values of $\gamma_{nm}$ and the functions $J_{nm}(\sigma)$. This expression can also be written as
\be
1 & =& \sum_{nm} P_{nm},
\label{eq:sumrule}
\ee
where the contribution of each eigenfunction is
\be
P_{nm} & \equiv & \sqrt{ \frac{3}{2} } \frac{16\pi^2R\Delta^2}{3kE}  (-1)^n \gamma_{nm}^{-1} \int d\sigma J_{nm}(\sigma).
\ee
These coefficients can have either sign depending on the value of $n$. The size of each contribution scales roughly as $0.5/(m+1/8)^{2/3}$, with a very weak dependence on $n$. This indicates the need for a large number of $n$ and $m$ to converge, i.e., it takes roughly ten times as many $m$ modes to reduce the size of $P_{nm}$ by a factor of ${\sim}$5. Numerical convergence would be obtained only when the size of additional $P_{nm}$ are negligible in the sum.

@@

We seek now to tabulate all eigenmodes $J_{nm}(\sigma)$ and eigenfrequencies $\gamma_{nm}$ for all spatial $n$ and for all eigenmodes $m$ at that given $n$. A resonant response in $J(n,\sigma,-i\gamma)$ is produced when the imaginary frequency is negative and has no real component. To measure the size of the response to detect where resonances occur, we sum the absolute value of $J(n,\sigma,-i\gamma)$ over $\sigma$ in array form. Call this response $f$, and use the index $j$ to represent the value of the response at discrete points over a range of $\gamma$. In places where $f_j > f_{j-1}$ and $f_j>f_{j+1}$, we have bounded a resonance that occurs in the interval $(\gamma_{j-1},\gamma_{j+1})$. We then call the ``solve'' step to hone in on the eigenvalue before continuing the sweep in $\gamma$. In the solve step, we evaluate the eigenfunctions at the points $(\gamma_{j-1},\gamma_{j},\gamma_{j+1})$ to obtain the responses $f_{j-1}$, $f_j$, and $f_{j+1}$. A ``guess'' at the correct eigenvalue $\gamma_{nm}$ can be calculated from
\be
\gamma_{\rm guess} &=& \frac{b\gamma_{j-1} - \gamma_{j+1}}{b - 1}
\ee
where
\be
b &=& \left(\frac{f_{j} - f_{j-1}}{f_{j} - f_{j+1}}\right)\left(\frac{\gamma_{j+1}-\gamma_{j}}{\gamma_{j-1}-\gamma_{j}}\right).
\ee
This is repeated by replacing former points $(\gamma_{j-1},\gamma_{j},\gamma_{j+1})$ with closer estimates while the size of the response grows as the resonance is approached. Given a sufficiently accurate eigenvalue $\gamma_{nm}$, we now find the corresponding eigenfunction $J_{nm}(\sigma)$. We can write down an approximate expression for the response near the resonances as
\be
J(n,\sigma,-i\gamma) & \simeq & \frac{J_{nm}(\sigma)}{\gamma_{nm}-\gamma} + C(\sigma),
\ee
where $C(\sigma)$ is weakly dependent on $\gamma$, and so we treat it as a constant. We can now solve for $J_{nm}(\sigma)$ by comparing two points near the resonance:
\be
J(n,\sigma,-i\gamma_1)  - J(n,\sigma,-i\gamma_2)& = &  \frac{J_{nm}(\sigma)}{\gamma_{nm}-\gamma_1} -  \frac{J_{nm}(\sigma)}{\gamma_{nm}-\gamma_2}
\ee
where $C(\sigma)$ has cancelled out in the difference. We can immediately solve as
\be
J_{nm}(\sigma) & \simeq & \frac{ J(n,\sigma,\gamma_1)  - J(n,\sigma,\gamma_2) }{ 1/(\gamma_{nm}-\gamma_1) - 1/(\gamma_{nm}-\gamma_2)}.
\ee
\begin{figure}
    \centering
    \includegraphics{Jsoln_n1_m100.pdf}
    \caption{Eigenfunctions $J_{nm}(x)$ for the lowest-order spatial eigenmode $n=1$, and a range of frequency eigenmodes up to $m=100$. Their oscillatory forms have varying amplitudes which sum to create the final form of the mean intensity. The largest contribution always comes from the ($n=1, m=1$) lowest-order eigenfunction.}
    \label{fig:jsoln}
\end{figure}
A set of eigenfunctions $J_{nm}(\sigma)$ are plotted in Fig. \ref{fig:jsoln} to illustrate their relative scales for different frequency eigenmodes $m$ at a fixed spatial eigenmode $n$. The overall scale of the $J_{nm}(\sigma)$ are set by the factor $E/(kR^3)$ with $E$ arbitrarily set to 1, $k = \tau_0 \sqrt{\pi} \Delta / R$, and $R = 10^{11}$ cm. With this normalization, an eigenfunction has typical size ${\sim} E a / \left(R^2 \Delta \right) = 10^{-37}$. Additional terms add smaller, faster-oscillating components that lead to higher accuracy upon summation with the lower-order terms. The oscillations of various modes must cancel perfectly to recreate the form of the solution in Eq. \ref{eq:Jrsigmat}, so many modes $m$ and $n$ are required for sufficient numerical precision in the final result. Too few (or imprecisely-calculated) modes creates ``ringing'' in the function that indicates the presence of imperfect cancellations in the positive and negative contributions of individual poles.

@@

\subsection{Frequency dependence}

Integrating $dE/dt d\nu$ from Eq. \ref{eq:dEdtdnu} over all times should give a distribution for the emitted frequencies. It is expected that this time-integrated spectrum of the response to an impulse will agree well with the solution for the $J=0$ steady-state spectrum (Eq. \ref{eq:H0surf}). Dividing by the energy $E$, we find the probability distribution
\be \label{eq:spectrum}
P(\nu) & = &  \frac{16\pi^2 R}{3k\phi E}  \sum_{nm} (-1)^n \gamma_{nm}^{-1} J_{nm}(\sigma).
\ee
Integrating over $\nu$ gives unity if the sum rule in Eq. \ref{eq:sumrule} is true. In Fig. \ref{fig:steadystate}, the resulting spectrum is shown for a sum up to $n=20$ and $m=500$ (labelled ``Time-integrated'') as compared with two analytic solutions: the steady-state $J=0$ solution with no time dependence (``Steady State'') and the expected result for summing \textit{only} spatial modes in the eigenfunction expansion with the time dependence ignored (``Partial Sum''). This graphic illustrates the physics modeled by each variable in the expansion. For example, the spatial modes contribute to the solutions' accuracy in the core of the line. If more spatial modes were calculated, the agreement between the lines extends further into the core. If additional frequency modes are calculated, faster-oscillating terms are incorporated into the sum over the imaginary frequency variable $\gamma$ which create more perfect cancellations with the lower-order terms, reducing the ``ringing'' seen in the time-integrated spectrum. Modeling the physics deep into the line core could have an impact on the accuracy of the escape time distribution, but this would only affect late times since core photons are trapped for a longer duration before escaping. This was the motivation for choosing a comparatively low number of spatial eigenmodes with respect to the number of frequency eigenmodes calculated.

@@

\subsection{Wait time distribution}

The distribution of the times it takes each photon to escape the sphere is calculated from Eq. \ref{eq:dEdtdnu} by integrating over all frequencies. We obtain
\be
P(t)  & = & \sqrt{\frac{3}{2}} \frac{16\pi^2 R \Delta^2 }{3kE}     \sum_{nm} (-1)^n  e^{-\gamma_{nm}t} \int d\sigma J_{nm}(\sigma) 
\nonumber \\ & = &  \sum_{nm} P_{nm} \gamma_{nm} e^{-\gamma_{nm}t},
\label{eq:waittime}
\ee
which is also normalized to unity.

The eigenfunction expansion method is an effective characterization of the time-to-escape, yet not a fast enough computational method to use as a distribution for sampling. However, the behavior of the solution is well-understood since the late-time distribution is simply an exponential falloff in photon escape probability. The $e$-folding of the function is the lowest-order damping rate, $\gamma_{00}$. This is shown in Fig. \ref{fig:escape_time}. The orange line, representing the exponential falloff piece, matches well with the late-time behavior past a point $ct/R = ct_{0}/R = (a\tau_0)^{1/3}$. The early-time behavior can be modeled well with a gamma distribution, provided an adequate fit. For sampling, however, these fitting coefficients would need to be derived as functions of the input parameters $\tau_0$, $a$, and $x_s$.

\begin{figure}
    \centering
    \includegraphics{waittime.pdf}
    \caption{A wait time distribution of escaping photons, which is fit using the lowest order eigenfrequency as a decay timescale for the exponential falloff. The solid blue curve is a full solution using a sum over $n$ and $m$ modes. A gamma function can also be found that characterizes the early time behavior well, but underestimates photons escaping at late times. \todo{Add optical depth and xinit info. Quantify what the error in the slope is as compared with the Monte Carlo. Redo the plot for the lower optical depths and do a comparison of the MC data for tau=$10^5$ and $10^6$ to see if the agreement is getting better. Could be that we're getting the wrong decay time because the photon is spending a long time in the doppler core and that's not represented in our solution. However, it could be a better and better approximation for larger tau. So if we see a worse disagreement for $10^5$ then we know it's actually converging. Could plot as a cumulative probability distribution --- the tail shouldn't be very important in the flux calculation. Do CDF for theory line as well --- integrate time dependence in exponent.}}
    \label{fig:escape_time}
\end{figure}

@@

The solution for mean intensity ($J$) and flux ($H = F/4\pi$) of photons near line-center frequency $\nu_0$ in a uniform sphere originally derived by \cite{2006ApJ...649...14D} has been generalized to allow a monochromatic source of photons with frequencies away from line center. However, this solution satisfies $J=0$ at the surface of the sphere, which is clearly unphysical as there could be no spectrum. A correction which allows the proper boundary condition $J=\sqrt{3}H$ to be satisfied comes from the introduction of the terms $J_{\rm bc}$ and $H_{\rm bc}$, which are solved using a continuous Fourier expansion in frequency. Enforcement of the correct boundary term leads naturally to expansion of the spatial dependence in terms of spherical Bessel functions. In practice, the continuous expansions are discretized and the Fourier coefficients solved for via a matrix equation. The resulting flux correction, $H_{\rm bc}$, scales as $(a\tau_0)^{-1/3}$ where $a$ is the damping parameter and $\tau_0$ is the line-center optical depth. Thus, for very large optical depths, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower optical depths. Additionally, since the Laplacian form of the differential equation is only attainable by approximating the line profile as $\phi \approx a/(\pi x^2)$, these solutions are invalid in areas where the Doppler core of the \lya line has a significant effect on the scattering of photons. Since the peak of the spectral energy distribution of escaping photons is $x_0 = (a\tau_0)^{1/3}$, calculations performed at lower optical depths are likely to be inaccurate due to the close proximity of the spectral peak and the Doppler core. By comparison with Monte Carlo simulations, it is seen that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions and correctly tracks changes in the spectrum of escaping photons as the monochromatic source frequency is altered, provided a sufficiently large optical depth.

The time-dependence of the problem can also be solved to

>>>>

\citet{1973MNRAS.162...43H, 1974MNRAS.166..373H} demonstrated that resonance-line radiation transfer in low-density, optically-thick media can be described by the Poisson equation. However, the accuracy of their solution is limited by the treatment of the boundary conditions. In their eigenfunction expansion, the spatial and frequency components of the differential equation are separated, yet the spatial eigenvalues are defined as being explicitly frequency-dependent. The resulting solution is therefore invalid on the basis that the differential equation and boundary conditions cannot simultaneously be satisfied. A more detailed discussion of the boundary condition is included in Section \ref{sec:steadystate}. \citet{1990ApJ...350..216N} follows this approach in an attempt to extend the result to media of intermediate optical depth, including the effects of scattering in the Doppler core of the line, but again does not use the correct boundary condition. This produces a divergent, inhomogeneous solution to the problem in 1D slab geometry. \citet{2006ApJ...649...14D} generalize the problem to spherical geometry, which is the same physical set up used here. Following this, \citet{2020arXiv200509692L} presents closed-form solutions for the radiation field in spherical geometry, utilizing the ``core skipping'' method discussed in \citet{2015MNRAS.449.4336S} to handle photon frequency redistribution within the Doppler core. This technique involves drawing the atom's velocity from a truncated Gaussian distribution, forcing photons back into the wing when they scatter into the core. This is motivated by the assumption that all core scattering results in zero spatial diffusion until the photon is redistributed back out into the wing where the mean free path becomes much larger and it can diffuse in space again. However, it is not clear whether this prescription is fully independent of the choice of velocity distribution.

@@

In order to understand how long it takes for the photons to escape the uniform sphere of gas we must reintroduce the time dependence of the diffusion equation, which was ignored in the steady-state calculations in Section \ref{sec:steadystate}. To obtain the radiative intensity $I=dE/(dAdtd\Omega d\nu)$ of a fluid that is dynamic on timescales shorter than the light-crossing time $t_{\rm lc} = R/c$, the time-dependent response to an impulse is found, i.e., a delta function in time. This allows the distribution of photon escape times (the ``wait time distribution'') to be characterized and fit for efficient sampling. For simplicity a $J=0$ boundary condition will be used in the following derivations, which is expected to be a decent approximation for $\tau_0 \gg 1$.

@@

Let us define the damping frequency to be $\gamma \equiv i\omega$, which is real-valued and positive for damped solutions. There are certain values of $\gamma$ for which a resonant response is produced which we will call $\gamma_{nm}$. Here, $m$ is an index that represents individual frequency eigenmodes, as opposed to $n$ which represents the spatial mode number. Near these frequencies, the solution will have the form 
\be
J(n,\sigma,-i\gamma) & \simeq \frac{ J_{nm}(\sigma) }{\gamma_{nm} - \gamma} + {\rm terms\ which\ vary\ slowly\ in\ \gamma}.
\ee
Obtaining the values of these eigenfrequencies and their corresponding eigenfunctions $J_{nm}(\sigma)$ allows us to obtain the contribution of this mode to the total solution via contour integration. 
\be
J(r,\sigma,t) & = &  -j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t}J(n,\sigma,-i\gamma)
\nonumber \\ & \simeq & j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t} \frac{ J_{nm}(\sigma) }{\gamma - \gamma_{nm}} 
\nonumber \\ & \simeq & j_0(\kappa_n r)  J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Here we have closed the integral along the imaginary $\gamma$ axis by a semi-circle at infinity, using the residue theorem to obtain the answer. The contribution from the semi-circle at infinity is zero. Summing over all spatial eigenmodes $n$ and over all temporal eigenmodes $m$ for a given $n$, we get the final form of the result
\be \label{eq:Jrsigmat}
J(r,\sigma,t) & = & \sum_{n=1}^\infty j_0(\kappa_n r)  \sum_m J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Beginning with Eq. \ref{eq:Jrsigmat}, we use the derivative
\be
\frac{dj_0(\kappa_n r)}{dr} \rfloor_R & =& \frac{d}{dr}  \rfloor_R \left( \frac{\sin(\kappa_n r)}{\kappa_n r} \right)
=  \left( \frac{\cos(\kappa_n R)}{R} - \frac{\sin(\kappa_n R)}{\kappa_n R^2} \right)\rfloor_R
\nonumber \\ & = & \frac{(-1)^n}{R},
\ee
to obtain the flux at the surface, which is
\be
F(R,\sigma,t) & =& - \frac{4\pi}{3k\phi} \frac{dJ(R,\sigma,t)}{dr} 
= - \frac{4\pi}{3k\phi R}  \sum_{nm} (-1)^n J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Multiplying by $4\pi R^2$ gives the energy per time per frequency
\be
\frac{dE}{dtd\nu} & = & - \frac{16\pi^2 R}{3k\phi}  \sum_{nm} (-1)^n J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\label{eq:dEdtdnu}
\ee
Integrating this over time yields a factor $-1/\gamma_{nm}$, and over $d\nu = \sqrt{3/2} \Delta^2 \phi d\sigma$, a ``sum rule" results
\be
1 & = &  \sqrt{ \frac{3}{2} } \frac{16\pi^2R\Delta^2}{3kE} \sum_{nm} (-1)^n \gamma_{nm}^{-1} \int d\sigma J_{nm}(\sigma).
\ee
This non-trivial expression provides a check on the values of $\gamma_{nm}$ and the functions $J_{nm}(\sigma)$. This expression can also be written as
\be
1 & =& \sum_{nm} P_{nm},
\label{eq:sumrule}
\ee
where the contribution of each eigenfunction is
\be \label{eq:pnmsoln}
P_{nm} & \equiv & \sqrt{ \frac{3}{2} } \frac{16\pi^2R\Delta^2}{3kE}  (-1)^n \gamma_{nm}^{-1} \int d\sigma J_{nm}(\sigma).
\ee
These coefficients can have either sign depending on the value of $n$. The size of each contribution scales roughly as $0.5/(m+1/8)^{2/3}$, with a very weak dependence on $n$. This indicates the need for a large number of $n$ and $m$ to converge, i.e., it takes roughly ten times as many $m$ modes to reduce the size of $P_{nm}$ by a factor of ${\sim}$5. Numerical convergence would be obtained only when the size of additional $P_{nm}$ are negligible in the sum.

@@

We seek now to calculate all eigenmodes $J_{nm}(\sigma)$ and eigenfrequencies $\gamma_{nm}$ for all spatial $n$ and for all eigenmodes $m$ at that given $n$. A resonant response in $J(n,\sigma,-i\gamma)$ is produced for special values of $\gamma \equiv \gamma_{nm}$. To measure the size of the response to detect where resonances occur, we sum the absolute value of $J(n,\sigma,-i\gamma)$ over $\sigma$ in array form. Call this response $f$, and use the index $j$ to represent the value of the response at discrete points over a range of $\gamma$. In places where $f_j > f_{j-1}$ and $f_j>f_{j+1}$, we have bounded a resonance that occurs in the interval $(\gamma_{j-1},\gamma_{j+1})$. We then call the ``solve'' step to hone in on the eigenvalue before continuing the sweep in $\gamma$. In the solve step, we evaluate the eigenfunctions at the points $(\gamma_{j-1},\gamma_{j},\gamma_{j+1})$ to obtain the responses $f_{j-1}$, $f_j$, and $f_{j+1}$. A ``guess'' at the correct eigenvalue $\gamma_{nm}$ can be calculated from
\be
\gamma_{\rm guess} &=& \frac{b\gamma_{j-1} - \gamma_{j+1}}{b - 1}
\ee
where
\be
b &=& \left(\frac{f_{j} - f_{j-1}}{f_{j} - f_{j+1}}\right)\left(\frac{\gamma_{j+1}-\gamma_{j}}{\gamma_{j-1}-\gamma_{j}}\right).
\ee
This is repeated by replacing former points $(\gamma_{j-1},\gamma_{j},\gamma_{j+1})$ with closer estimates while the size of the response grows as the resonance is approached. Given a sufficiently accurate eigenvalue $\gamma_{nm}$, we now find the corresponding eigenfunction $J_{nm}(\sigma)$. We can write down an approximate expression for the response near the resonances as
\be
J(n,\sigma,-i\gamma) & \simeq & \frac{J_{nm}(\sigma)}{\gamma_{nm}-\gamma} + C(\sigma),
\ee
where $C(\sigma)$ is weakly dependent on $\gamma$, and so we treat it as a constant. We can now solve for $J_{nm}(\sigma)$ by comparing two points near the resonance:
\be
J(n,\sigma,-i\gamma_1)  - J(n,\sigma,-i\gamma_2)& = &  \frac{J_{nm}(\sigma)}{\gamma_{nm}-\gamma_1} -  \frac{J_{nm}(\sigma)}{\gamma_{nm}-\gamma_2}
\ee
where $C(\sigma)$ has cancelled out in the difference. We can immediately solve as
\be
J_{nm}(\sigma) & \simeq & \frac{ J(n,\sigma,\gamma_1)  - J(n,\sigma,\gamma_2) }{ 1/(\gamma_{nm}-\gamma_1) - 1/(\gamma_{nm}-\gamma_2)}.
\ee
\begin{figure}
    \centering
    \includegraphics{Jsoln_n1_m100.pdf}
    \caption{Eigenfunctions $J_{nm}(x)$ for the lowest-order spatial eigenmode $n=1$, and a range of frequency eigenmodes up to $m=100$. Their oscillatory forms have varying amplitudes which sum to create the final form of the mean intensity. The largest contribution always comes from the ($n=1, m=1$) lowest-order eigenfunction.}
    \label{fig:jsoln}
\end{figure}
A set of eigenfunctions $J_{nm}(\sigma)$ are plotted in Fig. \ref{fig:jsoln} to illustrate their relative scales for different frequency eigenmodes $m$ at a fixed spatial eigenmode $n$. The overall scale of the $J_{nm}(\sigma)$ are set by the factor $E/(kR^3)$ with $E$ arbitrarily set to 1, $k = \tau_0 \sqrt{\pi} \Delta / R$, and $R = 10^{11}$ cm. With this normalization, an eigenfunction has typical size ${\sim} E a / \left(R^2 \Delta \right) = 10^{-37}$. Additional terms add smaller, faster-oscillating components that lead to higher accuracy upon summation with the lower-order terms. The oscillations of various modes must cancel perfectly to recreate the form of the solution in Eq. \ref{eq:Jrsigmat}, so many modes $m$ and $n$ are required for sufficient numerical precision in the final result. Too few (or imprecisely-calculated) modes creates ``ringing'' in the function that indicates the presence of imperfect cancellations in the positive and negative contributions of individual poles.

@@

\subsection{Comparison with Steady State and Monte Carlo}

Integrating $dE/dt d\nu$ from Eq. \ref{eq:dEdtdnu} over all times should give a distribution for the emitted frequencies. It is expected that this time-integrated spectrum of the response to an impulse will agree well with the solution for the $J=0$ steady-state spectrum (Eq. \ref{eq:H0surf}). Dividing by the energy $E$, we find the probability distribution
\be \label{eq:spectrum}
P(\nu) & = &  \frac{16\pi^2 R}{3k\phi E}  \sum_{nm} (-1)^n \gamma_{nm}^{-1} J_{nm}(\sigma).
\ee
Integrating over $\nu$ gives unity according to the sum rule in Eq. \ref{eq:sumrule}. In Fig. \ref{fig:steadystate}, the resulting spectrum is shown for a sum up to $n=20$ and $m=500$ (labelled ``Time-integrated'') as compared with two analytic solutions: the steady-state $J=0$ solution with no time dependence (``Steady State'') and the result for summing \textit{only} spatial modes in the eigenfunction expansion with the time dependence ignored (``Partial Sum''). This graphic illustrates the physics modeled by each variable in the expansion. For example, the spatial modes contribute to the solutions' accuracy in the core of the line. If more spatial modes were calculated, the agreement between the lines extends further into the core. If additional frequency modes are calculated, faster-oscillating terms are incorporated into the sum over the imaginary frequency variable $\gamma$ which create more perfect cancellations with the lower-order terms, reducing the ``ringing'' seen in the time-integrated spectrum. Extending the calculation deep into the line core by adding additional spatial modes could have an impact on the accuracy of the escape time distribution, but this would only affect the distribution at late times since core photons are trapped for a longer duration before escaping. This was the motivation for choosing a comparatively low number of spatial eigenmodes with respect to the number of frequency eigenmodes calculated. As a result of this choice, we expect that our semianalytic approach will underestimate the number of escaping photon at late times as compared with the Monte Carlo.

@@

The distribution of the times it takes each photon to escape the sphere is calculated from Eq. \ref{eq:dEdtdnu} by integrating over all frequencies. We obtain
\be
P(t)  & = & \sqrt{\frac{3}{2}} \frac{16\pi^2 R \Delta^2 }{3kE}     \sum_{nm} (-1)^n  e^{-\gamma_{nm}t} \int d\sigma J_{nm}(\sigma) 
\nonumber \\ & = &  \sum_{nm} P_{nm} \gamma_{nm} e^{-\gamma_{nm}t},
\label{eq:waittime}
\ee
which is also normalized to unity. For a sufficient number of spatial modes $n$ and frequency modes $m$, this sum approaches the ``true'' distribution of photon escape times as modeled by Monte Carlo simulations.

The eigenfunction expansion method is an effective characterization of the time-to-escape, yet not a fast enough computational method to use as a distribution for sampling. However, the behavior of the solution is well-understood since the late-time distribution is simply an exponential falloff in photon escape probability. The rate constant of the function is the lowest-order eigenfrequency, $\gamma_{00}$, and its scale is determined by the coefficient $P_{00}$ as in Eq. \ref{eq:pnmsoln}. Thus, an approximate ``fitting function'' that captures both the peak of the escape time distribution and the exponential falloff is
\be \label{eq:fitting_function}
P(t) = \exp{\left[-\left(\frac{t_{\rm diff}}{t}\right)^2\right]} \times \gamma_{00} P_{00} e^{-\gamma_{00}t}.
\ee
The first term represents the early-time distribution, which then transitions to an exponential falloff past a point $ct/R = ct_{\rm diff}/R = (a\tau_0)^{1/3}$, where $t_{\rm diff}$ is a characteristic diffusion timescale. This is shown in Fig. \ref{fig:escape_time} for two different line center optical depths. It is seen that the fitting function matches well with the late-time behavior of the eigenfunction solution, indicating that the tail of the escape time distribution can be modeled effectively by simple exponential decay. The only remaining discrepancy between the tail of the distribution and the Monte Carlo data is due to core scattering, which is not accurately modeled by the eigenfunction solution as previously discussed. A cumulative distribution function can be created from Eq. \ref{eq:fitting_function} using a simple rejection method, which allows the function to be sampled rapidly.

\begin{figure}
    \centering
    \includegraphics{fitting_formula.pdf}
    \caption{A wait time distribution of escaping photons for two different line center optical depths $\tau_0$, which is fit using Eq. \ref{eq:fitting_function}. A sum over 20 spatial eigenmodes and 500 frequency eigenmodes is labeled ``Eigenfunctions''. All calculations were performed with a monochromatic source of photons at line center ($x_s = 0.0$).}
    \label{fig:escape_time}
\end{figure}

The effect of many scatterings in the Doppler core affects the tail of the escape-time distribution, since photons that become trapped here will take much longer to escape on average. This explains the deficit of photons in the semi-analytic escape time distributions at late times, i.e., the rate constant for the exponential falloff is overestimated slightly as compared with the Monte Carlo. The error in this rate constant is a function of line center optical depth, since the effect from the Doppler core is greatest when it extends into the area of the spectrum where photons most readily escape. In Fig. \ref{fig:tau_scaling}, it is shown that the rate constant of exponential decay converges with the lowest-order eigenfrequency $\gamma_{00}$ at sufficiently high optical depths, following a $t\propto(a\tau_0)^{1/3}$ scaling. At lower optical depths $\tau_0$, the effects of core scattering are most important, leading to a larger discrepancy in the characteristic escape timescale. Here, the Monte Carlo accurately characterizes the photons which scatter in the core many times before escaping, while the semi-analytic solution does not capture this behavior as it uses only the Lorentzian piece of the line profile and does not use enough spatial modes to accurately model the frequency regime near line center. However, as the optical depth grows, the effect of core scattering becomes smaller and the approximations hold much better, agreeing well with the expected analytic scaling for the rate constant.

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{tau_scaling.pdf}
    \caption{Characteristic timescales as a function of optical depth. The reciprocal of the lowest-order eigenfrequency, $\gamma_{00}$, and a fit of the exponential falloff for the Monte Carlo data are shown to converge to a $t \propto (a\tau_0)^{1/3}$ scaling at large optical depths. The Monte Carlo points on this figure are obtained by fitting only the exponential piece of the escape time distribution to obtain the rate constant.}
    \label{fig:tau_scaling}
\end{figure}

@@

The solution for mean intensity ($J$) and flux ($H = F/4\pi$) of photons near line-center frequency $\nu_0$ in a uniform sphere originally derived by \cite{2006ApJ...649...14D} has been generalized to allow a monochromatic source of photons with frequencies away from line center. A correction to the $J=0$ surface boundary condition which allows the proper condition $J=\sqrt{3}H$ to be satisfied comes from the introduction of the terms $J_{\rm bc}$ and $H_{\rm bc}$, which are solved using a continuous Fourier expansion in frequency. Enforcement of the correct boundary term leads naturally to expansion of the spatial dependence in terms of spherical Bessel functions. In practice, the continuous expansions are discretized and the Fourier coefficients solved for via a matrix equation. The resulting flux correction, $H_{\rm bc}$, scales as $(a\tau_0)^{-1/3}$ where $a$ is the damping parameter and $\tau_0$ is the line-center optical depth. Thus, for very large optical depths, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower optical depths. Additionally, since the Laplacian form of the differential equation is only attainable by approximating the line profile as $\phi \approx a/(\pi x^2)$, these solutions are invalid in areas where the Doppler core of the \lya line has a significant effect on the scattering of photons. Since the peak of the spectral energy distribution of escaping photons is $x_0 = (a\tau_0)^{1/3}$, calculations performed at lower optical depths are likely to be inaccurate due to the close proximity of the spectral peak and the Doppler core. By comparison with Monte Carlo simulations, it is seen that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions and correctly tracks changes in the spectrum of escaping photons as the monochromatic source frequency is altered, provided a sufficiently large optical depth.
